\Large \bf DRAFT: Direct-style process creation on Linux

### 2020-05-27 research/paper/paper.tex
paper

\titleformat*{\subsection}{\normalsize\bfseries}
\titleformat*{\subsubsection}{\normalsize\bfseries}
\titlespacing{\subsection}{0pt}{1ex}{0.5ex}
\titlespacing{\subsubsection}{0pt}{0.5ex}{0ex}
\lstset{
language=Python,
basicstyle=\small\ttfamily,
keywordstyle=,
stringstyle=\color{red},
frame=tb,
columns=fullflexible,
showstringspaces=false,
belowskip=0pt,
}

@@

In these examples and in this paper,
we'll have multiple processes,
all of which are under the control of a single program with a single flow of control.
Because of the absence of multiple concurrent control flows,
these examples could be executed interactively by a human operator,
typing them in line by line in a REPL.
We'll always be working from the perspective of this single program;
never from the perspective of a child process, because (until we call \texttt{exec} and give up control)
there is no meaningful program running in the child processes.
\subsection{Abstraction through file descriptor passing}\label{fd_abstraction}
In Linux (and Unix in general),
a program can pass file descriptors from the parent process to a child process
by using file descriptor inheritance.\cite{tlpi}
When a process is created, all open file descriptors are copied from the parent process to the child process.
The child process can then make independent use of the file descriptors.
We'll discuss this important feature further in section \ref{fdtables}.

@@

\lstinputlisting[float,language=Python,label={lst:fds},caption={Passing down FDs}]
{examples/fds/direct.py}
In Listing \ref{lst:fds},
we first open a file with read-write permission in the local process.
Then we create a child process,
which inherits all file descriptors from its parent.
We indicate that we want to operate on the child's inherited copy of the file descriptor with \verb|child.inherit_fd|.
\texttt{inherit\_fd} performs no system calls,
it just updates bookkeeping to return a new handle for the inherited file descriptor;
we'll discuss it in more depth in section \ref{handles}.
Then we disable \texttt{CLOEXEC} on the child's file descriptor,
which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|;
this ensures that the file descriptor will be usable by the new program after we call \texttt{exec}.
Finally, we execute a new program in the child process,
passing the file descriptor number as an argument.
The new program will be able to use the file descriptor we opened through that file descriptor number.
\subsection{Non-shared-memory concurrency}
Processes run concurrently,
which enables modularity
and allows exploiting the parallelism of the underlying hardware.
Since processes don't share memory,
they can provide a less complex parallel programming environment
than shared-memory thread-based approaches.\cite{threads}

@@

\lstinputlisting[float,language=Python,label={lst:fuse},caption={Shared file descriptor tables}]
{examples/fuse/direct.py}
In listing \ref{lst:fuse},
we first create a child process \verb|ns_child| in a new mount namespace with \verb|CLONE_NEWNS|,
and this time also pass \verb|CLONE_FILES|,
which causes the file descriptor table to be shared between the parent process and the child process.\cite{clone}
We create another child from \verb|ns_child|
and use it to \texttt{exec} a FUSE filesystem,
which will appear only in the mount namespace of \verb|ns_child| and its descendents.
Then we can open FUSE files in \verb|ns_child|
and use those files in \texttt{local},
through the shared file descriptor table.
We call \texttt{use\_fd} which, like \texttt{inherit\_fd},
updates bookkeeping to create a new handle for the file descriptor existing in the shared fd table;
we'll discuss \texttt{use\_fd} in section \ref{handles}.

@@

We use this to create processes running the syscall server.
After this,
the parent process can be begin to call system calls in the child process.
Most system calls work as normal;
the new child process can be modified freely through \texttt{chdir}, \texttt{dup2}, and other system calls.
From the system calls related to process creation,
only \texttt{execve} and \texttt{unshare} need substantial further attention.
\subsection{\texttt{exec}}\label{execve}
\texttt{execve} is unusual and requires careful design,
because when it is successful, it does not return.
We need a way to determine if \texttt{execve} is successful;
naively waiting for a response to the syscall request may leave us waiting forever.

@@

Some other Unix-like systems natively provide the ability to wait for a child's \texttt{execve};
our implementation would be simplified, and the ambiguity eliminated,
if we had this ability on Linux.\cite{freebsd_kqueue}
One approach would be to add a new \texttt{clone} flag to
opt-in to receiving \texttt{WEXECED} events through \texttt{waitid}.
A new \texttt{waitid} flag alone is not sufficient,
since, to integrate this feature into an event loop,
it's necessary to receive \texttt{SIGCHLD} signals or readability notifications on a pidfd
when the \texttt{WEXECED} event happens.
\subsection{Managing file descriptor tables}\label{fdtables}
\texttt{clone} (without \verb|CLONE_FILES|) and \texttt{fork}
deal with file descriptor tables in the same way:
The new child process has a new file descriptor table,
containing copies of all the file descriptors existing in the parent at the time of the system call.
This is known as file descriptor inheritance.

As demonstrated in section \ref{shared_fd_table},
when the \verb|CLONE_FILES| flag is passed to \texttt{clone},
the parent process and the child process will share a single file descriptor table,
so changes to the file descriptor table made in one process will affect the other.

The file descriptor table can then be unshared by calling \verb|unshare(CLONE_FILES)| or \texttt{execve}.
After either system call, the process will have a new file descriptor table
which is no longer shared with any other process.
All the file descriptors existing in the old table at the time of the system call
will be copied into the new file descriptor table,
except for, in the case of \texttt{execve}, file descriptors with the \texttt{CLOEXEC} flag set.
\subsubsection{Tracking file descriptors through handles}\label{handles}
After \texttt{clone}, \verb|unshare(CLONE_FILES)|, or \texttt{execve}
has copied all the file descriptors from some old table into a new table,
we need to decide which file descriptors we want to keep open in the new table,
and which file descriptors should be closed.

In general, if a certain file descriptor is in use by our program,
through a certain process,
that file descriptor should stay open in that process's file descriptor table.

To achieve this, file descriptors are used through garbage-collected handles;
each handle is associated with a process and contains a specific file descriptor number.
A handle is ``valid''
when that file descriptor number refers to an open file descriptor in that process.
A handle is ``live''
when the handle is in use by the program, which we can determine
from Python's garbage collection system.
We then keep open exactly those file descriptors which are necessary to keep all live handles, valid.

When a process changes file descriptor table through \verb|unshare(CLONE_FILES)|,
we need to make sure all the live file descriptor handles associated with that process continue to be valid.
This is straightforward:
we enumerate all the handles associated with the process,
and don't close the file descriptors that those handles refer to.

Likewise, if a process changes file descriptor table through \texttt{execve}
in a way that allows the program to keep control of the process,
we make sure to keep all the file descriptors with live handles open.
This requires careful management of \texttt{CLOEXEC}.

When a child process is created,
there are no file descriptor handles associated with it.
We can make file descriptor handles by opening file descriptors in the child,
or by creating new handles for file descriptors opened in other processes.

We can use the \verb|inherit_fd| function to create handles for inherited file descriptors.
\verb|inherit_fd| takes a file descriptor handle associated with the parent process,
and, if that file descriptor was open at the time the child process was created,
returns a new handle associated with the child process
for the child process's copy of that file descriptor.
Once the inherited file descriptors are closed as described in section \ref{cloexec},
\verb|inherit_fd| can no longer be used.

If the process shares its file descriptor table with any other processes,
we can use the \verb|use_fd| function to create handles for file descriptors
in the shared file descriptor table.
\verb|use_fd| takes as arguments a process $A$
and a file descriptor handle associated with any process sharing process $A$'s file descriptor table;
it returns a new handle associated with process $A$ for that file descriptor.
\subsubsection{Closing file descriptors after inheritance}\label{cloexec}
After creating a new file descriptor table,
and determining which file descriptors should be kept open,
we must promptly close other file descriptors.
Leaving unwanted file descriptors open in the new table wastes resources,
and can also cause bugs.
For example, it's common practice to close the write end of a pipe
and expect an EOF on the read end;
if the write end is copied into a new file descriptor table between creating and closing,
and the write end is never closed in the new table,
the read end will never get an EOF.

% TODO should definitely cite this
However, we can't simply close all other file descriptors,
because we must preserve the semantics of implicit file descriptor inheritance.
The possibility of implicit inheritance of file descriptors is an old Unix feature,
which is useful in a wide variety of situations,
in much the same way as environment variables, which are also implicitly inherited.
As with environment variables,
implicitly inherited file descriptors allow a resource to be passed down a process hierarchy
without intervening programs being aware;
this is a form of dynamic scope, which, though notorious, is sometimes useful.\cite{implicit_params}

Fortunately, file descriptors which should be implicitly inherited are marked:
they have \texttt{CLOEXEC} turned off.
For implicit inheritance to function, \texttt{CLOEXEC} must not be set;
otherwise, the file descriptor will be closed, rather than inherited,
the next time the process calls \texttt{execve}.
And since \texttt{execve} can be called at any time in a general purpose program,
any library that does not want its file descriptors to be inherited across \texttt{execve}
must keep \texttt{CLOEXEC} turned on at all times.
So, at any time, we know that the file descriptors with \texttt{CLOEXEC} off
are exactly the file descriptors which should be implicitly inherited.

Therefore, we simply need to close all file descriptors that have \texttt{CLOEXEC} set,
and which are not referenced by a live file descriptor handle;
we refer to this operation as \verb|close_cloexec|.

When we call \texttt{unshare(CLONE\_FILES)},
we perform \texttt{close\_cloexec} automatically in userspace.
For \texttt{clone}, we perform \verb|close_cloexec| only on user request,
so that the user can control when \verb|inherit_fd| stops working.
Even if the user never requests that unused file descriptors be closed,
as long as they call \texttt{execve}, the table will be cleaned at that point.
In both of these cases,
we would benefit from a new system call which takes a list of file descriptors
and closes all \texttt{CLOEXEC} file descriptors except for those on the list;
this would replace our userspace implementation of \verb|close_cloexec|.

For \texttt{execve}, \verb|close_cloexec| is, in some sense, done automatically for us by the kernel;
except we don't get the chance to preserve live file descriptors handles,
other than by unsetting \texttt{CLOEXEC} on them before calling \texttt{execve}.
Again an extension to the kernel would be useful:
A new argument to \texttt{execve} which allows specifying a list of file descriptors
which should be inherited across \texttt{execve} even if \texttt{CLOEXEC} is set on them.
This new argument would allow programs to avoid the pattern
of unsetting \texttt{CLOEXEC}, calling \texttt{execve}, and then immediately re-setting \texttt{CLOEXEC},
just to inherit a file descriptor across a single \texttt{execve}.
Avoiding that pattern would primarily be useful
when the process calling \texttt{execve} has a shared file descriptor table;
in that case, unsetting \texttt{CLOEXEC} before \texttt{execve} is not possible,
because some other process might also call \texttt{execve}
and erroneously inherit the file descriptor from the shared file descriptor table.
\section{Evaluation}\label{evaluation}
Our goal is a powerful approach to process creation that is easy to use correctly.
Performance of our implementation is secondary,
but the overhead of our implementation relative to alternative approaches
must be low.

@@

Several other aspects of our implementation would also be improved by new features in the Linux kernel,
as discussed in sections \ref{execve} and \ref{cloexec}.
\subsection{Portability to other Unix systems}
Other non-Linux systems
could adopt the techniques of this paper
to provide direct-style process creation.
Currently, our focus is on Linux,
but others may wish to explore porting these techniques to other operating systems.
% TODO maybe we could get rid of this paragraph?
\subsection{Large scale open source usage}
We would like to open source libraries built on top of direct-style process creation.
From our experience using such libraries, discussed in section \ref{realworld},
we believe this would make it easier to develop complex systems involving processes.
\section{Conclusions}\label{conclusions}
We have introduced direct-style process creation on Linux for the first time,
through \texttt{rsyscall}.
This style of process creation is common on academic operating systems,
but was previously not usable on Linux.
We've provided a number of examples which demonstrate the usefulness
of direct-style process creation in section \ref{examples}.
Our current implementation of \texttt{rsyscall} has excellent support for Python,
works entirely in userspace,
and is immediately deployable on today's Linux systems.
\texttt{rsyscall} is open source and available from \githuburl{}.

>>>>

% \titleformat*{\subsection}{\normalsize\bfseries}
% \titleformat*{\subsubsection}{\normalsize\bfseries}
% \titlespacing{\subsection}{0pt}{1ex}{0.5ex}
% \titlespacing{\subsubsection}{0pt}{0.5ex}{0ex}
\lstset{
language=Python,
basicstyle=\small\ttfamily,
keywordstyle=,
stringstyle=\color{red},
frame=tb,
columns=fullflexible,
showstringspaces=false,
belowskip=0pt,
}

@@

In these examples and in this paper,
we will have multiple processes,
all of which are under the control of a single program with a single flow of control.
Because of the absence of multiple concurrent control flows,
these examples could be executed interactively by a human operator,
typing them in line by line in a REPL.
We'll always be working from the perspective of this single program;
never from the perspective of a child process, because (until we call \texttt{exec} and give up control)
there is no meaningful program running in the child processes.
We believe this absence of concurrency simplifies reasoning.
\subsection{Abstraction through FD passing}\label{fd_abstraction}
In Linux (and Unix in general),
a program can pass file descriptors (FDs) from the parent process to a child process
by using file descriptor inheritance.\cite{tlpi}
When a process is created, all open file descriptors are copied from the parent process to the child process.
The child process can then make independent use of the file descriptors.
We'll discuss this important feature further in section \ref{inheritance}.

@@

\lstinputlisting[float,language=Python,label={lst:fds},caption={Passing down FDs}]
{examples/fds/direct.py}
In Listing \ref{lst:fds},
we first open a file with read-write permission in the local process.
Then we create a child process,
which inherits all file descriptors from its parent.
We indicate that we want to operate on the child's inherited copy of the file descriptor with \verb|child.inherit_fd|.
\texttt{inherit\_fd} performs no system calls,
it just updates bookkeeping to return a new handle for the inherited file descriptor;
we'll discuss it in more depth in section \ref{inheritance}.
Then we disable \texttt{CLOEXEC} on the child's file descriptor,
which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|;
this ensures that the file descriptor will be usable by the new program after we call \texttt{exec}.
Finally, we execute a new program in the child process,
passing the file descriptor number as an argument.
The new program will be able to use the file descriptor we opened through that file descriptor number.
\subsection{Non-shared-memory concurrency}
Processes run concurrently,
which enables modularity
and allows exploiting the parallelism of the underlying hardware.
Since processes don't share memory,
they can provide a less complex parallel programming environment
than shared-memory thread-based approaches.\cite{threads}

@@

\lstinputlisting[float,language=Python,label={lst:fuse},caption={Shared file descriptor tables}]
{examples/fuse/direct.py}
In listing \ref{lst:fuse},
we first create a child process \verb|ns_child| in a new mount namespace with \verb|CLONE_NEWNS|,
and this time also pass \verb|CLONE_FILES|,
which causes the file descriptor table to be shared between the parent process and the child process.\cite{clone}
We create another child from \verb|ns_child|
and use it to \texttt{exec} a FUSE filesystem,
which will appear only in the mount namespace of \verb|ns_child| and its descendents.
Then we can open FUSE files in \verb|ns_child|
and use those files in \texttt{local},
through the shared file descriptor table.
We call \texttt{use\_fd} which, like \texttt{inherit\_fd},
updates bookkeeping to create a new handle for the file descriptor existing in the shared fd table;
we'll discuss \texttt{use\_fd} in section \ref{inheritance}.

@@

We use this to create processes running the syscall server.
After this,
the parent process can be begin to call system calls in the child process.
Most system calls work as normal;
the new child process can be modified freely through \texttt{chdir}, \texttt{dup2}, and other system calls.
From the system calls related to process creation,
only \texttt{execve} needs substantial further attention.
\subsection{\texttt{exec}}\label{execve}
\texttt{execve} is unusual and requires careful design,
because when it is successful, it does not return.
We need a way to determine if \texttt{execve} is successful;
naively waiting for a response to the syscall request may leave us waiting forever.

@@

Some other Unix-like systems natively provide the ability to wait for a child's \texttt{execve};
our implementation would be simplified, and the ambiguity eliminated,
if we had this ability on Linux.\cite{freebsd_kqueue}
One approach would be to add a new \texttt{clone} flag to
opt-in to receiving \texttt{WEXECED} events through \texttt{waitid}.
A new \texttt{waitid} flag alone is not sufficient,
since, to integrate this feature into an event loop,
it's necessary to receive \texttt{SIGCHLD} signals or readability notifications on a pidfd
when the \texttt{WEXECED} event happens.
\subsection{Handling file descriptors}\label{fd_handles}
In Unix, there are many types of resources maintained inside the kernel,
such as file descriptors or memory mappings,
which are referred to by identifiers which are valid only within a certain context\cite{tlpi}.
For example, file descriptors are referred to by file descriptor numbers
which are valid only in a certain file descriptor table;
memory and memory mappings are referred to by memory addresses
which are valid only in a certain address space;
the same pattern is followed for a number of other resources.
Each process has one file descriptor table, one address space, and one of each of the other contexts.
Two or more processes can share individual contexts;
for example, two processes might share an address space, but not a file descriptor table.

The core issues here are the same for all such resources,
so we'll focus on file descriptors.
For concision and clarity, we'll abbreviate ``file descriptor'' as ``FD'' in this section.

A typical POSIX program operates within only one process,
and so only needs to concern itself with one of each kind of context.
To identify and operate on an FD, it can use just an FD number,
which is valid within the global implicit process's FD table.
That's sufficient to make syscalls involving that FD.

A program using \texttt{rsyscall}, however,
operates in multiple processes,
and has to deal with resources across multiple contexts.
To identify an FD,
it needs not just a FD number,
but also some kind of identifier for an FD table.
To then operate on that FD,
it needs a capability for a process with that FD table,
so that it can make syscalls in that process.

To manage this, we add a new concept alongside FD, FD number, and FD table:
The FD handle.
An FD handle is a pair of an FD number and a process capability.
This both precisely identifies an FD, and allows operating on it.
We say an FD handle is associated with a process
if that FD handle contains that process capability.

All system calls which would otherwise return FD numbers or take FD numbers as an argument,
now instead return FD handles or take FD handles as an argument.
For convenience, as shown in section \ref{examples} for \texttt{fcntl},
helper methods on the FD handle object are provided for many system calls dealing with a single FD;
these methods simply call the corresponding underlying system call
using the process capability contained in the FD handle.

In this way, we can work with FDs without the ambiguity of dealing with raw FD numbers in a multi-process context.
\subsubsection{File descriptor inheritance and table sharing}\label{inheritance}
The picture is complicated by the need to support some FD behaviors which implicitly cross between processes.
FDs are copied between processes at process creation time through FD inheritance,
and two processes can share the same FDs through a shared FD table.

FD inheritance is a behavior exhibited by \texttt{clone} (without \verb|CLONE_FILES|) and by \texttt{fork}.
When these system calls are called,
they create a new child process with a new FD table.
The new FD table contains copies of all the FDs existing in the parent process's FD table at the time of the system call,
at the same FD numbers.

FD table sharing occurs when \texttt{clone} \textit{with} \texttt{CLONE\_FILES} is called;
this creates a new child process that shares its FD table with its parent.
As a result, either process can operate on any FD in the table,
through the same FD numbers.

These behaviors are unusual in that they operate on all open FDs at once,
rather than individual FDs explicitly denoted by system call arguments.
There are other ways to pass FDs between processes,
such as \texttt{SCM\_RIGHTS}\cite{scm_rights},
but those operate on individual FDs through explicit arguments to system calls.
One or more FD handles are explicitly passed to a system call in one process,
and one or more FD handles are explicitly returned from another system call in another process.
The behaviors of FD inheritance and FD table sharing happen for all FDs, implicitly,
and so they must be dealt with differently to be fully supported.

\texttt{rsyscall} supports both behaviors in the same way,
by allowing us to create new FD handles on-demand
referring to already-open FDs created by FD inheritance or FD table sharing;
by creating new FD handles in new processes, we can call syscalls in those processes on those FDs.
\texttt{rsyscall} provides two functions
\texttt{inherit\_fd} and \texttt{use\_fd}, which support FD inheritance and FD table sharing, respectively.
Both functions take as arguments a process capability $A$ and an FD handle $H$ associated with some other process $B$,
and return a new FD handle associated with process $A$.
For \texttt{inherit\_fd},
this new handle refers to an \textit{inherited copy} of the FD which $H$ refers to;
\texttt{inherit\_fd} will fail if $B$ is not $A$'s parent process,
or if $H$ was not open at the time $A$ was created.
For \texttt{use\_fd},
this new handle refers to the \textit{same} FD that $H$ refers to, through the shared FD table;
\texttt{use\_fd} will fail if $A$ and $B$ don't share the same FD table.
\section{Evaluation}\label{evaluation}
Our goal is a powerful approach to process creation that is easy to use correctly.
Performance of our implementation is secondary,
but the overhead of our implementation relative to alternative approaches
must be low.

@@

Other aspects of our implementation would also be improved by new features in the Linux kernel;
one examples is discussed in section \ref{execve}.
\subsection{Portability to other Unix systems}
Other non-Linux systems
could adopt the techniques of this paper
to provide direct-style process creation.
Currently, our focus is on Linux,
but others may wish to explore porting these techniques to other operating systems.
% TODO maybe we could get rid of this paragraph?
\subsection{Large scale open source usage}
We would like to open source libraries built on top of direct-style process creation.
From our experience using such libraries, discussed in section \ref{realworld},
we believe this would make it easier to develop complex systems involving processes.
\section{Conclusions}\label{conclusions}
We have introduced direct-style process creation on Linux for the first time,
through \texttt{rsyscall}.
This style of process creation is common on academic operating systems,
but was previously not usable on Linux.
We've provided a number of examples which demonstrate the usefulness
of direct-style process creation in section \ref{examples}.
Our current implementation of \texttt{rsyscall} has excellent support for Python,
works entirely in userspace,
and is immediately deployable on today's Linux systems.
\texttt{rsyscall} is open source and available from \githuburl{}.
